Type Classes for Mathematics in Type Theory

### 2010-07-01 jar.tex
References for manifest fields, more precise about adjunctions

Manifest fields~\cite{DBLP:conf/types/Luo08} have been proposed, and implemented in the Matita system, to address exactly this problem. We hope to convince the reader that type system extensions of this nature, designed to mitigate particular symptoms of the bundled approach, are less elegant than a solution (described in the next section) that avoids the problem altogether by using predicate-like type classes in place of bundled records.

@@

Adjunctions can be defined in different ways. A nice symmetric definition is the following:
\begin{lstlisting}
  Class Adjunction (φ: Π `(F c ⟶ d), (c ⟶ G d)): Prop :=
    { adjunction_left_functor: Functor F _
    ; adjunction_right_functor: Functor G _
    ; natural_left `(f: d ⟶ d') c: (fmap G f ◎) ∘ φ$$ = φ (c:=c) ∘ (f ◎)
    ; natural_right `(f: c' ⟶ c) d: (◎ f) ∘ φ (d:=d) = φ$$ ∘ (◎ fmap F f) }.
\end{lstlisting}
An alternative definition is the following:
\begin{lstlisting}
  Class AltAdjunction (η: id ⇛ G ∘ F) (φ: Π `(f: c ⟶ G d), F c ⟶ d): Prop :=
    { alt_adjunction_natural_unit: NaturalTransformation η
    ; alt_adjunction_factor: Π `(f: c ⟶ G d),
        is_sole ((f =) ∘ (◎ η$$ c) ∘ fmap G) (φ f) }.
\end{lstlisting}
Formalizing the (nontrivial) proof that these two definitions are equivalent provides a nice test for our definitions. Preliminary results indicate that we can very concisely and closely follow the textbook proof. % todo: ref?

@@

For the present example, we define an ad-hoc term language for monoids
\begin{lstlisting}
  Inductive Expr (V: Type) := Mult (a b: Expr V) | One | Var (v: V).
\end{lstlisting}
The expression type is parameterized over the set of variable indices. Hence, we diverge from~\cite{Hints}, which uses \lstinline|nat| for variables, thereby introducing bound problems and a need for dummy variables.

>>>>

Manifest fields~\cite{Pollack:2002} have been proposed to address exactly this problem. In fact, a semblance has been implemented in the Matita system~\cite{sacerdoti2008working}. We hope to convince the reader that type system extensions of this nature, designed to mitigate particular symptoms of the bundled approach, are less elegant than a solution (described in the next section) that avoids the problem altogether by using predicate-like type classes in place of bundled records.

@@

Adjunctions can be defined in different ways. A nice symmetric definition is the following:
\begin{lstlisting}
  Class Adjunction (φ: Π `(F c ⟶ d), (c ⟶ G d)): Prop :=
    { adjunction_left_functor: Functor F _
    ; adjunction_right_functor: Functor G _
    ; natural_left `(f: d ⟶ d') c: (fmap G f ◎) ∘ φ$$ = φ (c:=c) ∘ (f ◎)
    ; natural_right `(f: c' ⟶ c) d: (◎ f) ∘ φ (d:=d) = φ$$ ∘ (◎ fmap F f) }.
\end{lstlisting}
An alternative definition is the following:
\begin{lstlisting}
  Class AltAdjunction (η: id ⇛ G ∘ F) (φ: Π `(f: c ⟶ G d), F c ⟶ d): Prop :=
    { alt_adjunction_natural_unit: NaturalTransformation η
    ; alt_adjunction_factor: Π `(f: c ⟶ G d),
        is_sole ((f =) ∘ (◎ η$$ c) ∘ fmap G) (φ f) }.
\end{lstlisting}
Formalizing the (nontrivial) proof that these two definitions are equivalent provides a nice test for our definitions. As a first step, we have constructed the unit and co-unit of the adjunction, thus proving MacLane's Theorem 1. We have concisely and closely followed his proof~\cite{CatWork}. % todo: ref?

@@

For the present example, we define an ad-hoc term language for monoids
\begin{lstlisting}
  Inductive Expr (V: Type) := Mult (a b: Expr V) | One | Var (v: V).
\end{lstlisting}
The expression type is parameterized over the set of variable indices. Below we use an implicitly defined heap of such variables. Hence, we diverge from~\cite{Hints}, which uses \lstinline|nat| for variables, thereby introducing bound problems and a need for dummy variables.
